Statically Typed String Sanitation Inside a Python

### 2014-08-13 sanitation-psp14/sanitation-psp14-short.tex

To be practical, adopting a specialized type system like this should not require the adoption of a new programming language. Instead, we favor extensible type systems: new type system fragments like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.  We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
% Security-oriented programming languages do not enjoy wide adoption. Conversely, library and framework-based approaches to security are pervasive. Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly. In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks. To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms. We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library. We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
\end{abstract}

@@

To avoid this problem, the program must \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats as safely. Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging. Note that this function is not idempotent, so it should only be called once.

@@

In this paper, we present a static type system that tracks the regular language a string belongs to. For example, the output of \verb|sanitize| will be in the regular language described by the regular expression above. We write this as $\mathcal{L}\{r\}$. We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}. These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner. The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types (e.g. at the edges of the system, where user input has not yet been validated). %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

@@

% Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library. This approach is \emph{composable} in the sense that it is a conservative extension. This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.

@@

The premier operation for manipulating strings in $\lambdas$ is string substitution, 
which is a familiar operation to any programmer who has used regular expressions.
The replacement operation replaces all instances of a pattern in one string with
another string; for instance, $\lsubst{a|b}{a}{c} = c$.
In order to computer the type resulting from aubstitution, we also need to compute
the result of replacing on language with another inside a given language.
Finally, just for convienance, we provide a coerce operation. The introduction of
coercion requires handling of runtime errors.

@@

A brief outline of this section follows:
\begin{itemize}
  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
   \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
\end{itemize}

@@

@@

\addtocounter{figure}{4}
\begin{figure*}\label{fig:trans}

@@

\newpage
\ 
\newpage
\ 
\newpage

@@

@@

\section{Implementation as a Library in {Atlang}}\label{atlang}
\begin{figure}
\begin{lstlisting}
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') # TODO: is this right?
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

\begin{figure}
\begin{lstlisting}
output of failed compilation
\end{lstlisting}
\caption{Output of failed compilation.}
\end{figure}
wwwww
Here, we use the argument annotation syntax introduced in Python 3 (a similar syntax is available for Python 2.6+, not shown).

\section{Related Work}\label{related}
The input sanitation problem is well-understood. There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks. In this section, we explain how our approach to the input sanitation problem differs from each of these approaches. More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming lanugage design.

@@

We are also unaware of any extensible programming languages which emphasize applications to security concerns.

Incorporating regular expressions into the type system is not novel. The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions. 
Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
We differ from this and related work in at least three ways:
\begin{itemize}
  \item Our system is defined within an extensible type system.
  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schemas.
  \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
\end{itemize}

In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
\section{Conclusion}

An implementation of a similar system as a type system extension in the programming language Ace is discussed in \cite{fulton12} and \cite{fulton13}.
The implementation only supports a special case of replacement where unsafe substrings are simply stripped.
In practice, most libraries and frameworks escape command sequences instead of stripping characters; therefore, we plan to extend our implementation with
support for the replace operation as described in this paper..

>>>>

To be practical, adopting a specialized type system like this should not require the adoption of a new programming language. Instead, we favor extensible type systems: new type system fragments like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.  We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.

% Security-oriented programming languages do not enjoy wide adoption. Conversely, library and framework-based approaches to security are pervasive. Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly. In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks. To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms. We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library. We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
\end{abstract}

@@

To avoid this problem, the program must \emph{sanitize} user input. For example, in this case, the developer (or, more often, a framework) might define a function \verb|sanitize| that prepends double quotes (and existing backslashes)  with a backslash, which SQL treats safely. Guaranteeing that user input has already been sanitized before it is used to construct a command is challenging. Note that this function is not idempotent, so it should only be called once.

@@

In this paper, we present a static type system that tracks the regular language a string belongs to. For example, the output of \verb|sanitize| will be in the regular language described by the regular expression above. We write this as $\mathcal{L}\{r\}$. We take advantage of closure and decidability properties of regular languages to support a number of useful operations on values of such \emph{regular string types}, including \emph{replacement} of substrings matching a given regular expression. These make it possible to implement sanitation protocols like the one just described in an essentially conventional manner. The result is a system where the fact that a string has been {correctly} sanitized is manifest in its type. Missing calls to sanitization functions are detected statically, and, importantly, so are \emph{incorrectly implemented sanitization functions} (i.e. these functions need not be trusted). These guarantees require run-time checks only when going from less precise to more precise types (e.g. at the edges of the system, where user input has not yet been validated). %implementing and statically checking input sanitation techniques. Our solution suggests a more general approach to the integration of security concerns into programming language design. This approach is characterized by \emph{composable} type system extensions which \emph{complement} existing and well-understood solutions with compile-time checks.

@@

% Summarily, we present a simple type system extension which ensures the absence of input sanitation vulnerabilities by statically checking input sanitation algorithms which use an underlying regular expression library. This approach is \emph{composable} in the sense that it is a conservative extension. This approach is also \emph{complementary} to existing input sanitation techniques which use string replacement for input sanitation.
% Our type system for regular string types, $\lambdas$, is defined as an exntension to a simply-typed lambda calculus.
% Operations on strings are given types according to corresponding, type-level
% computations on regular expressions. A concatenation operation is defined in 
% order to demonstrate this principle, but the premier operation of our system
% is \emph{replacement}, which captures a rich set of input sanitation techniques
% including escaped characters and encodings. In addition to defining this system,
% we present a correctness-preserving compilation of $\lambdas$ into an underlying
% language with ordinary strings as a regular expression library.

% In addition to this theoretical contribution, we also explain how our extension
% may be implemented in the Ace Programming Language, which is itself an extension
% of Python -- a popular web programming language.

% Summarily, we present a simple type system extension which ensures the absence
% of input sanitation vulnerabilities by statically checking the correctness and 
% correct use of input sanitation algorithms. This approach is \emph{composable}
% in the sense that it is a conservative extension. This approach is \emph{complementary}
% to existing input sanitation techniques that use regular expressions for input sanitation.

@@

The premier operation for manipulating strings in $\lambdas$ is string substitution, 
which is a familiar operation to any programmer who has used regular expressions.
The replacement operation replaces all instances of a pattern in one string with
another string; for instance, $\lsubst{a|b}{a}{c} = c$.
In order to compute the type resulting from substitution, we also need to compute
the result of replacing one language with another inside a given language.
Finally, just for convenience, we provide a coerce operation. The introduction of
coercion requires handling of runtime errors.

@@

A brief outline of this section follows:
\begin{itemize}
  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$. Grammar follows immediately at the top of page 4.
  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
   \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
  \item In \S 2.3 we prove that translation preserves the correctness result about $\lambdas$.
\end{itemize}

@@

% \begin{figure}[h]
% \begin{grammar}

% <$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
% $\str$ \alt $\regex$

% <P> ::= $\lambda x . e$ \hfill target terms \alt
%   $e e$ \alt
%   $\tstr{s}$ \alt
%   $\rx{r}$ \alt
%   $\tconcat{P}{P}$ \alt
%   $\preplace{P}{P}{P}$ \alt
%   $\tcheck{P}{P}$ 
%   \end{grammar}
% \caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
% \label{fig:lcsSyntax}
% \end{figure}

@@

\addtocounter{figure}{4}
\begin{figure*}[t]\label{fig:trans}

@@

\newpage
\ 
\newpage
\ 
\newpage
\

% Begin Grammar 
% Removes the < > from grammar productions.
\renewcommand{\grammarlabel}[2]{#1\hfill#2}

% \begin{figure}[t]
% \begin{grammar}
% <r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

% \caption{Regular expressions over the alphabet $\Sigma$.}
% \label{fig:regex}
% \end{grammar}
% \end{figure}

% \begin{figure}[h]
% \begin{grammar}

% <$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
% $\stringin{r}$

% <S> ::= 
%       $\lambda x . e$ \hfill source terms \alt
%       $e e$ \alt
%       $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
%       $\rsconcat{S}{S}$ \alt
%       $\sisubst{r}{S}{S}$ \alt
%       $\coerce{r}{S}$
% \caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
% \end{grammar}
% \end{figure}

\subsection{Definition of $\lambdas$}

The $\lambdas$ system extends the simply-typed lambda calculus with regular
expression types; an explanation of significant typing rule follows:
\begin{itemize}
  \item Rule S-T-Concat states that the type of concatenated strings is 
    obtained by concatenating the regular expressions for each string.
  \item Rule S-T-Replace defines the type of $\textsf{rreplace}$. The expression $\textsf{rreplace}$
    evaluates to the result of substituting every string matching $r$ in $s_1$
    with $s_2$. This operation corresponds to the \textsf{str_replace} function of PHP.
    The type of these expressions is defined in terms of an extra-linguistic \textsf{lreplace}
    function, which is defined later in this section.
    The expression $\lreplace{r}{r_1}{r_2}$ is obtained by
    starting with $r_1$, and replacing any subexpression matching $r$ with $r_2$.
  \item Rule S-T-Coerce allows coercion; however, note that this coercion cannot
    invalidate our safety property because we ensure appropraite checks are always
    inserted wherever coercions are used.
\end{itemize}

\subsection{Definition of $\lambdap$}

The system $\lambdap$ is a straight-forward extension of a simply typed lambda calculus
with a string type and a regular expression tpye. We include two operations which are available in the regular expression
library of any modern programming language. The check operation ensures that an expression
recognizes a string, and the replace operation is string replacement.

\subsection{Definition of Translation}

The translation from $\lambdas$ to $\lambdap$ is defined in figure 5.
The coercion cases are most interesting. If the safety of coercion in manifest in the
types of the expressions, then no runtime check is inserted.
If the safety of coercion is not manifest in the types, then a check is inserted.

In practice, the type of a replacement rarely matches a specification.
Therefore, it is convienant in an implementation to always insert the appropriate
coercion, and then only raise type errors when an automatically inserted coercion
actually requires the insertion of a runtime check. Alternatively, this policy may 
be codified in the type system itself using subtyping \cite{fulton12}.

@@

The main theorem of this paper is Theorem \ref{thm:main},
which establishes that Theorem \ref{thm:scorrect} is preserved under translation
into the target language $\lambdap$.

Establishing this result requires an additional theorem establishing a relationship between canonical forms for the string fragments of $\lambdas$ and $\lambdap$.

@@

\section{Implementation in {Atlang}}\label{atlang}
\begin{figure}\label{fig:example}
\begin{lstlisting}
@fn
def sanitize(s : string_in[r'.*']):
  return (s.replace(r'"', '&quot;') 
           .replace(r'<', '&lt;')
           .replace(r'>', '&gt;'))

@@

\begin{figure}
\begin{lstlisting}
output of failed compilation
\end{lstlisting}
\caption{Output of failed compilation.}
\end{figure}

\subsection{An Example}

Figure \ref{fig:example} demosntrates the use of string types in atlang.
The sanitize function takes an arbitrary string and returns a string without
double quotes or left and right brackets. In this example, we use HTML
escape sequences.

The main function receives user input and passes this input to a 
sanitize function, which replaces all double quotes and brackets with HTML
escape sequences.

The result of applying sanitize to input is appended to two functions which
construct a safe query and safe output.
The arguments to the result and output construction functions constitute \emph{specifications}.
In the case of \textsf{results_query}, this specification ensures that user input is always interpreted
as a string literal by the SQL server.
In the case of \textsf{results_div}, this specification ensures that user input does 
not contain any HTML tags, which is a conservative but effective policy for preventing
XSS attacks.

Note that ${\sf input}$ does not actually meet these specifications without additional machinery.
The type of ${\sf input}$ is quite large
and does not actually equal the specified domains of the query or output construction
methods. This mismatch is common -- in fact, nearly universal. Therefore, 
our implementation includes a simple subtyping relation between regular expression
types. 
This subtyping relation is justified theoretically by the fact that language inclusion is
decidable; see \cite{fulton12} for a formal definition of the subtyping relation.
Additionally, our extension remains composable because subtyping is defined on a type-by-type basis;
see \cite{fulton13} for a discussion of subtyping in Atlang (referred to there as Ace).

\subsection{Implementation of the Regular Expression Type}

We implemented a variation on the type system presented in this paper. The only
significant difference is that we only support replacements where $s_2$ is the empty
string. Therefore, our implementation respects the system presented in section 2
only modulo the definition of \textsf{lreplace}.

Atlang translates programs using type definitions, which may extend both the
static and dynamic semantics of the language. New types are defined as Python
classes; figure \ref{fig:impl} contains the source code of our implementation.

The \textsf{string_in} type has an indexing regular expression \texttt{idx}.
Our translation is defined by the \textsf{trans_} methods while the \textsf{syn_}
methods define our type checker. Atlang defers type checking and translation
to these methods whenever an expression of type \textsf{string_in} is encountered.

\subsection{Related Work and Alternative Approaches}

The input sanitation problem is well-understood. There exist a large number of techniques and technologies, proposed by both practitioners and researchers, for preventing injection-style attacks. In this section, we explain how our approach to the input sanitation problem differs from each of these approaches. More important than these differences, however, is our more general assertion that language extensibility is a promising approach toward consideration of security goals in programming language design.

@@

The Wyvern programming language provides a general framework for composing language extensions \cite{tsl}\cite{wyvern}.
Our work identifies one particular extension, and is therefore complementary to Wyvern and related
work on extensible programming languages.
We are also unaware of any extensible programming languages which emphasize applications to security concerns.

Incorporating regular expressions into the type system is not novel. The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schema using regular expressions. 
Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
We differ from this and related work in at least three ways:
\begin{itemize}
  \item Our system is defined within an extensible type system.
  \item We demonstrate that regular expression types are applicable to the web security domain, whereas previous work on regular expression types focused on XML schema.
  \item Although our static replacement operation is definable in some languages with regular expression types, we are the first to expose this operation and connect the semantics of regular language replacement with the semantics of string substitution via a type safety and compilation correctness argument.
\end{itemize}

In conclusion, our contribution is a type system, implemented within an extensible type system, for checking the correctness of input sanitation algorithms.
